Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-11-02 TACO-paper-v2/coffee.tex
edits

The time required to execute the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells to obtain a satisfactory approximation
of the PDE, possibly of the order of trillions, as in
\citeN{Rossinelli2013}. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centers have a strict time
limit in which to produce a forecast---60 minutes in the case of the
UK Met Office. Producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper.

@@

In particular, we address the well-known problem of optimizing the
local assembly phase of the finite element method
\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
can be responsible for a significant fraction of the overall
computation run-time, often in the range 30-60$\%$. With respect to
these studies, we propose a novel set of composable code
transformations targeting, for the first time, cross-loop arithmetic
intensity, with emphasis on instruction-level parallelism, redundant
computation and register locality. How such transformations generalize
to domains other than finite element local assembly will also be
discussed.

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in the
discretized domain. In this work, we focus on relatively low order finite
element methods, in which an assembly kernel's working set is usually small
enough to fit the L1 cache of traditional CPU architectures. Low order methods are by no means exotic: they
are employed in a wide variety of fields, including climate and ocean
modeling, computational fluid dynamics, and structural mechanics. The
efficient assembly of high order methods such as the spectral element
method \cite{SPENCER} requires a significantly different loop nest structure.
High order methods are therefore excluded from our study.

We aim to maximize our impact on the platforms that are realistically used
for finite element applications, so we target conventional CPU architectures
rather than GPUs. 
The key limiting factor to the execution on GPUs is the stringent
memory requirements.  Only relatively small problems fit in a GPU
memory, and support for distributed GPU execution in general purpose
finite element frameworks is minimal. There has been some research on
adapting local assembly to GPUs, which we mention in
Section~\ref{sec:related-work}, although it differs from ours in
several ways, including: (i) not optimizing for cross-loop arithmetic
intensity (the goal is rather effective multi-thread parallelization),
and (ii) not relying on automated code generation from a
domain-specific language (explained next); moreover, they have done
testing only on order 1 methods, while we focus on the order range
1--4. In addition, our code transformations would drastically impact
the GPU parallelization strategy, for example by increasing a thread's
working set.  For all these reasons, a study on extending our work to
GPU architectures is beyond the scope of this paper. In
Section~\ref{sec:generality}, however, we provide some intuitions
about this research direction.

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific, compute
intensive expression evaluates a two dimensional array, representing
the result of local assembly in an element of the discretized
domain. With such kernel and working set structures, traditional
transformations for cache locality (e.g., blocking) are not helpful;
instead we focus on aspects like the minimization of floating-point
operations, register locality and instruction-level parallelism,
especially vectorization.

Achieving high performance on CPUs is non-trivial. The complexity of
the mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine
an optimal sequence of transformations that is successfully applicable
to different problems (equations). Loop trip counts are typically
small and can vary significantly, which further exacerbates the
issue. A compiler-based approach is, therefore, the only reasonable
option to obtain close-to-peak performance in a wide range of
different local assembly kernels. Optimizations like padding,
generalized loop-invariant code motion, vector-register tiling, and
expression splitting, as well as their composition, are essential, but
their support in state-of-the-art polyhedral and vendor compilers, if
any, is very limited. BLAS routines could theoretically be employed,
although a fairly complicated control and data flow analysis would be
required to automate identification and extraction of matrix-matrix
multiplies. In addition, as detailed in
Section~\ref{sec:perf-eval-blas}, the small dimension of the matrices
involved and the potential loss in data locality can limit or
eliminate the performance gain of this approach.

@@

Summarizing, the contributions of this paper are:
\begin{enumerate}
\item An optimization strategy for finite element local assembly that
  exploits domain knowledge and goes beyond the limits of both vendor
  and research compilers.
\item The design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item A systematic analysis using a suite of examples of real-world
  importance that is evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \item A full application evaluation to demonstrate that the optimization
 strategy greatly impacts the whole finite element problem.
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}
Our contributions also include a theoretical study involving
\begin{enumerate}
 \setcounter{enumi}{\theenumTemp}
 \item An analysis of the generality of the proposed code transformations
 and an investigation of their applicability to different computational
 domains.
\end{enumerate}
The paper is organized as follows. In Section~\ref{sec:background} we
provide some background on local assembly, show code generated by
Firedrake and emphasize the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. The generality of our optimization strategy
and the applicability to other domains is discussed in Section~\ref{sec:generality}.
Related work is described in Section~\ref{sec:related-work}. Finally,
Section~\ref{sec:conclusions} reviews our contributions in the light
of our results, and identifies priorities for future work.

@@

Local assembly is the computation of contributions of a specific cell in
the discretized domain to the linear system which yields the PDE
solution. The process consists of numerically evaluating
problem-specific integrals to produce a matrix and a vector
\cite{quadrature1,fluidity_manual_v4}, whose sizes depend on the order
of the method. This operation is applied to all cells in the discretized
domain. In this work we focus on local matrices, or \emph{element
matrices}, which are more costly to compute than element vectors.

Given a finite element description of the input problem, expressed
through the domain-specific Unified Form Language (UFL)~\cite{ufl},
Firedrake employs the FEniCS form compiler (FFC)~\cite{FFC-Compiler} to
generate an abstract syntax tree of a kernel implementing assembly using
numerical quadrature. This kernel can be applied to any
element in the mesh, which follows from a mathematical property of the
finite element method. The evaluation of a local matrix can be
reduced to integration on a fixed \emph{reference} element --- a special
element that does not belong to the domain --- after a suitable change
of coordinates. Firedrake triggers the compilation of an assembly
kernel using an available vendor compiler, and manages its (parallel)
execution over all elements in the mesh. As already explained, the
subject of this paper is to enhance this execution model by adding an
optimization stage prior to the generation of C code.

@@

The structure of a local assembly kernel is shown in
Figure~\ref{code:general-structure}. The inputs are a zero-initialized
two dimensional array used to store the element matrix, the element's
coordinates in the discretized domain, and coefficient fields, for
instance indicating the values of velocity or pressure in the
element. The output is the evaluated element matrix. The kernel body
can be logically split into three parts:
\begin{enumerate}
  \item Calculation of the Jacobian matrix, its determinant and its
    inverse required for the aforementioned change of coordinates from
    the reference element to the one being computed.
  % FIXME: this rather hand-wavy and not technically correct
  \item Definition of basis functions used to interpolate fields at the
    quadrature points in the element. The choice of basis functions is
    expressed in UFL directly by users. In the generated code, they are
    represented as global read-only two dimensional arrays (i.e., using
    \texttt{static const} in C) of double precision floats.
  \item Evaluation of the element matrix in an affine loop nest, in which
    the integration is performed.
\end{enumerate}
Table~\ref{table:map-name-letters} shows the variable names
we will use in the upcoming code snippets to refer to the various
kernel objects.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, has short trip counts (in the range 3--15), and the computation
reduces to a summation of a few products involving basis functions. An
example is provided in Listing~\ref{code:helmholtz}, which shows the
assembly kernel for a Helmholtz problem using Lagrange basis functions
on 2D elements with polynomial order $p=1$. In other scenarios, for
instance when solving the Burgers equation, the number of arrays
involved in the computation of the element matrix can be much larger.
The assembly code is given in Listing~\ref{code:burgers} and contains 14
different arrays that are accessed, where the same array can be referenced
multiple times within the same expression. This may also require the
evaluation of constants in outer loops (called $F$ in the code) to act
as scaling factors of arrays. Trip counts grow proportionately to the
order of the method and arrays may be block-sparse. In addition to a
larger number of operations, more complex cases like the Burgers
equation are characterized by high register pressure.

@@

Despite the infinite variety of assembly kernels which Firedrake can
generate, it is still possible to identify common domain-specific
traits that can be exploited for effective code transformations and
SIMD vectorization. These include: 1) memory accesses along the three
loop dimensions are always unit stride; 2) the \texttt{j} and \texttt{k}
loops are interchangeable, whereas interchanges involving the $i$ loop
require pre-computation of values (e.g. the $F$ values in Burgers) and
introduction of temporary arrays, as explained in
Section~\ref{sec:code-transf}; 3) depending on the problem being
solved, the \texttt{j} and \texttt{k} loops could iterate over the
same iteration space; 4) most of the sub-expressions on the right hand
side of the element matrix computation depend on just two loops
(either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}). In
Section~\ref{sec:code-transf} we show how to exploit these
observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to
all finite element problems that can be formulated in Firedrake.
Their generality and the potential for applicability in other domains is also discussed in
Section~\ref{sec:generality}.

As already emphasized, the variations in the structure of
mathematical expressions and in loop trip counts, although typically
limited to the order of tens of iterations, render the optimization process
challenging, requiring distinct sets of transformations to bring performance
closest to the machine peak in different problems. For example, the Burgers problem in Listing~\ref{code:burgers},
given the large number of arrays accessed, suffers from high register pressure,
whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does
not; this intuitively suggests that the two problems require a
different treatment, based on an in-depth analysis of both data and
iteration spaces. Furthermore, domain knowledge enables
transformations that a general-purpose compiler could not apply,
making the optimization space even larger. In this context, our goal
is to understand the relationship between distinct code
transformations, their impact on cross-loop arithmetic intensity, and to what
extent their composability is effective in a class of problems and
architectures.

@@

Data alignment is enforced in two steps. Firstly, all arrays are
allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then,
two dimensional arrays are padded by rounding the number of columns to
the nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the
original size of a basis function array is 3$\times$3 and
$\mbox{\texttt{VL}}=4$ (e.g. AVX processor, with 256-bit long vector
registers and 64-bit double-precision floats). In this case, a padded
version of the array will have size 3$\times$4. The compiler is
explicitly told about data alignment using suitable pragmas; for
example, in the case of the Intel compiler, the annotation
\texttt{$\#$pragma vector aligned} is added before the loop (as shown in
later figures) to inform that all of the memory accesses in the loop
body will be properly aligned. This allows the compiler to issue aligned
load and store instructions, which are notably faster than unaligned
ones.

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From the inspection of the codes in Listings~\ref{code:helmholtz}
and~\ref{code:burgers}, it can be noticed that the computation of $A$
involves evaluating many sub-expressions which only depend on two
iteration variables. Since symbols in most of these sub-expressions
are read-only variables, there is ample space for loop-invariant code
motion. Vendor compilers apply this technique, although not in the
systematic way we need for our assembly kernels. We want to overcome
two deficiencies that both Intel and GNU compilers exhibit. First,
they only identify sub-expressions that are invariant with respect to
the innermost loop. This is an issue for sub-expressions depending on
\texttt{i}-\texttt{k}, which are not automatically lifted in the loop
order \texttt{ijk}. Second, the hoisted code is scalar and therefore not
subjected to auto-vectorization.

@@

\KwSty{void} helmholtz(\KwSty{double} A[3][4], \KwSty{double} **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~\KwSty{double} LI$\_$0[4] ALIGN;\\
~~~~\KwSty{double} LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code for the Helmholtz problem in
  Listing~\ref{code:helmholtz} after application of padding, data
  alignment, and \emph{licm}, for an AVX architecture. In this
  example, sub-expressions invariant to \texttt{j} are identical to
  those invariant to \texttt{k}, so they can be precomputed once in
  the $r$ loop.}
\label{code:helmholtz-licm}
\end{algorithm}

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
\label{sec:op-vect}
One notable problem of assembly kernels concerns register allocation
and register locality. The critical situation occurs when loop trip
counts and the variables accessed are such that the vector-register
pressure is high. Since the kernel's working set fits the L1 cache, it
is particularly important to optimize register management. Standard
optimizations, such as loop interchange, unroll, and unroll-and-jam,
can be employed to deal with this problem. In COFFEE, these
optimizations are supported either by means of explicit code
transformations (interchange, unroll-and-jam) or indirectly by
delegation to the compiler through standard pragmas (unroll). Tiling
at the level of vector registers is an additional feature of
COFFEE. Based on the observation that the evaluation of the element
matrix can be reduced to a summation of outer products along the
\texttt{j} and \texttt{k} dimensions, a model-driven vector-register
tiling strategy can be implemented. If we consider the code snippet in
Listing~\ref{code:helmholtz-licm} and we ignore the presence of the
operation \texttt{det*W3[i]}, the computation of the element matrix is
abstractly expressible as
\begin{equation}
\scriptsize
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,2
\end{equation}
where $B$ is the set of all basis functions (or temporary variables,
e.g., \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$
are generic problem-dependent subsets. Regardless of the specific
input problem, by abstracting from the presence of all variables
independent of both \texttt{j} and \texttt{k}, the element matrix
computation is always reducible to this form.
Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries
($j,k=0,...,3$) of the element matrix using just 2 vector registers,
which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B''
\vert = 1$. Values in a register are shuffled each time a product is
performed. Standard compiler auto-vectorization for both GNU and Intel
compilers, instead, executes 4 broadcast operations (i.e., ``splat'' of
a value over all of the register locations) along the outer dimension to
perform the calculation. In addition to incurring a larger number of
cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers
to perform the same 16 evaluations when unroll-and-jam is used, with $f$
being the unroll-and-jam factor.

@@

The compiler applies an ordered sequence of optimization steps to the
ASTs received from FFC. Application of \emph{licm} must precede padding
and data alignment, due to the introduction of temporary arrays. These
transformations are always performed because they are likely to improve
the run-time performance, as demonstrated by our results in Section~\ref{sec:analysis-code-transf}.
Based on a cost model (described later), loop interchange, \emph{split},
and \emph{op-vect} may be introduced. Their implementation is centered
on analysis and manipulation of the kernel AST. When \emph{op-vect}
is selected, the compiler outputs AVX or AVX-512 intrinsics code. Any
possible corner cases are handled: for example, if \emph{op-vect} is to
be applied, but the size of the iteration space is not a multiple of the vector
length, then a remainder loop, amenable to auto-vectorization, is inserted.

@@

\subsubsection{Impact of Generalized Loop-invariant Code Motion}
\label{licm-impact}

>>>>

The time required to execute the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells to obtain a satisfactory approximation of the
PDE, possibly of the order of trillions, as in
\citeN{Rossinelli2013}. For example, it has been well established that
mesh resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centers have a strict time
limit in which to produce a forecast---60 minutes in the case of the
UK Met Office. Producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Computational cost is a dominant problem in computational
science simulations, especially for those based on finite elements,
that are the subject of this paper.

@@

In particular, we address the well-known problem of optimizing the
local assembly phase of the finite element method
\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
can be responsible for a significant fraction of the overall
computation run-time, often in the range 30-60$\%$. With respect to
these studies, we propose a novel set of composable code
transformations targeting, for the first time, cross-loop arithmetic
intensity, with emphasis on instruction-level parallelism, redundant
computation and register locality. Also, we discuss 
how such transformations generalize
to domains other than finite element local assembly.

During the assembly phase, the solution of the PDE is approximated by
executing a problem-specific kernel over all cells, or elements, in
the discretized domain. In this work, we focus on relatively low order
finite element methods, in which an assembly kernel's working set is
usually small enough to fit in the L1 cache of traditional CPU
architectures. Low order methods are by no means exotic: they are
employed in a wide variety of fields, including climate and ocean
modeling, computational fluid dynamics and structural mechanics. The
efficient assembly of high order methods such as the spectral element
method \cite{SPENCER} requires a significantly different loop nest
structure.  High order methods are therefore excluded from our study.

We aim to maximize our impact on the platforms that are realistically
used for finite element applications, so we target conventional CPU
architectures rather than GPUs.  The key limiting factor for the
execution on GPUs is the stringent memory requirements.  Only
relatively small problems fit in GPU memory, and support for
distributed GPU execution in general-purpose finite element frameworks
is minimal. There has been some research on adapting local assembly to
GPUs, which we mention in Section~\ref{sec:related-work}; these works 
differ from ours in several ways, including: (i) not optimizing for
cross-loop arithmetic intensity (the goal is rather effective
multi-thread parallelization), and (ii) not relying on automated code
generation from a domain-specific language (explained next); moreover,
they have done testing only on order 1 methods, while we focus on the
order range 1--4. In addition, our code transformations would
drastically impact the GPU parallelization strategy, for example by
increasing a thread's working set.  For all these reasons, a study on
extending our work to GPU architectures is beyond the scope of this
paper. In Section~\ref{sec:generality}, however, we provide some
intuitions about this research direction.

An assembly kernel is characterized by the presence of an affine,
often non-perfect loop nest, in which individual loops are rather
small: their trip count rarely exceeds 30, and may be as low as 3 for
low order methods. In the innermost loop, a problem-specific
compute-intensive expression evaluates a two dimensional array,
representing the result of local assembly in an element of the
discretized domain. With such kernel and working set structures,
traditional transformations for cache locality (e.g., blocking) are
not helpful; instead, we focus on aspects such as the optimization of
floating-point operations, register locality and instruction-level
parallelism, especially vectorization.

Achieving high performance on CPUs is non-trivial. The complexity of
the mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine
an optimal sequence of transformations that is successfully applicable
to different problems (equations). Loop trip counts are typically
small and can vary significantly and this further exacerbates the
issue. A compiler-based approach is, therefore, the only reasonable
option to obtain close-to-peak performance in a wide range of
different local assembly kernels. Optimizations like padding,
generalized loop-invariant code motion, vector-register tiling, and
expression splitting, as well as their composition, are essential, but
their support in state-of-the-art vendor and polyhedral compilers, if
any, is very limited. BLAS routines could theoretically be employed,
although a fairly complicated control and data flow analysis would be
required to automate identification and extraction of matrix-matrix
multiplications. In addition, as detailed in
Section~\ref{sec:perf-eval-blas}, the small dimension of the matrices
involved and the potential loss in data locality can limit or
eliminate the performance gain of this approach.

@@

Summarizing, the contributions of this paper are:
\begin{enumerate}
\item An optimization strategy for finite element local assembly that
  exploits domain knowledge and goes beyond the limits of both vendor
  and research compilers.
\item The design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item A systematic analysis using a suite of examples of real-world
  importance that is evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \item A full application evaluation to demonstrate that the optimization
 strategy greatly impacts the whole finite element problem.
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}
Our contributions also include a theoretical study involving
\begin{enumerate}
 \setcounter{enumi}{\theenumTemp}
 \item An analysis of the generality of the proposed code transformations
 and an investigation of their applicability to different computational
 domains.
\end{enumerate}
The paper is organized as follows. In Section~\ref{sec:background}, we
provide some background on local assembly, show code generated by
Firedrake and emphasize the critical computational
aspects. Section~\ref{sec:code-transf} describes the various code
transformations, highlighting when and how domain knowledge has been
exploited. The design and implementation of our compiler is discussed
in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
shows performance results. The generality of our optimization strategy
and the applicability to other domains is discussed in Section~\ref{sec:generality}.
Related work is described in Section~\ref{sec:related-work}. Finally,
Section~\ref{sec:conclusions} reviews our contributions in the light
of our results and identifies priorities for future work.

@@

Local assembly is the computation of contributions of a specific cell
in the discretized domain to the linear system that yields the PDE
solution. The process consists of numerically evaluating
problem-specific integrals to produce a matrix and a vector
\cite{quadrature1,fluidity_manual_v4}, whose sizes depend on the order
of the method. This operation is applied to all cells in the
discretized domain. In this work we focus on local matrices or
\emph{element matrices}, which are more costly to compute than element
vectors.

Given a finite element description of the input problem expressed
through the domain-specific Unified Form Language (UFL)~\cite{ufl},
Firedrake employs the FEniCS form compiler (FFC)~\cite{FFC-Compiler} to
generate an abstract syntax tree of a kernel implementing assembly using
numerical quadrature. This kernel can be applied to any
element in the mesh, which follows from a mathematical property of the
finite element method. The evaluation of a local matrix can be
reduced to integration on a fixed \emph{reference} element---a special
element that does not belong to the domain---after a suitable change
of coordinates. Firedrake triggers the compilation of an assembly
kernel using an available vendor compiler, and manages its parallel
execution over all elements in the mesh. As already explained, the
subject of this paper is to enhance this execution model by adding an
optimization stage prior to the generation of C code.

@@

The structure of a local assembly kernel is shown in
Figure~\ref{code:general-structure}. The inputs are a zero-initialized
two dimensional array used to store the element matrix, the element's
coordinates in the discretized domain and coefficient fields, for
instance indicating the values of velocity or pressure in the
element. The output is the evaluated element matrix. The kernel body
can be logically split into three parts.
\begin{enumerate}
  \item Calculation of the Jacobian matrix, its determinant and its
    inverse required for the aforementioned change of coordinates from
    the reference element to the one being computed.
  % FIXME: this rather hand-wavy and not technically correct
  \item Definition of basis functions used to interpolate fields at the
    quadrature points in the element. The choice of basis functions is
    expressed in UFL directly by users. In the generated code, they are
    represented as global read-only two dimensional arrays (i.e., using
    \texttt{static const} in C) of double precision floats.
  \item Evaluation of the element matrix in an affine loop nest in which
    the integration is performed.
\end{enumerate}
Table~\ref{table:map-name-letters} shows the variable names
we will use in the upcoming code snippets to refer to the various
kernel objects.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, has short trip counts (in the range 3--15), and the computation
reduces to a summation of a few products involving basis functions. An
example is provided in Listing~\ref{code:helmholtz} that shows the
assembly kernel for a Helmholtz problem using Lagrange basis functions
on 2D elements with polynomial order $p=1$. In other scenarios, for
instance when solving the Burgers equation, the number of arrays
involved in the computation of the element matrix can be much larger.
The assembly code is given in Listing~\ref{code:burgers} and contains 14
different arrays that are accessed, where the same array can be referenced
multiple times within the same expression. This may also require the
evaluation of constants in outer loops (called $F$ in the code) to act
as scaling factors of arrays. Trip counts grow proportionately to the
order of the method and arrays may be block-sparse. In addition to a
larger number of operations, more complex cases like the Burgers
equation are characterized by high register pressure.

@@

Despite the infinite variety of assembly kernels that Firedrake can
generate, it is still possible to identify common domain-specific
traits that can be exploited for effective code transformations and
SIMD vectorization. These include: 1) memory accesses along the three
loop dimensions are always unit stride; 2) the \texttt{j} and \texttt{k}
loops are interchangeable, whereas interchanges involving the $i$ loop
require pre-computation of values (e.g. the $F$ values in Burgers) and
introduction of temporary arrays, as explained in
Section~\ref{sec:code-transf}; 3) depending on the problem being
solved, the \texttt{j} and \texttt{k} loops could iterate over the
same iteration space; 4) most of the sub-expressions on the right hand
side of the element matrix computation depend on just two loops
(either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}). In
Section~\ref{sec:code-transf} we show how to exploit these
observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to
all finite element problems that can be formulated in Firedrake.
Their generality and the potential for applicability in other domains
is discussed in Section~\ref{sec:generality}.

As already emphasized, the variations in the structure of mathematical
expressions and in loop trip counts, although typically limited to
tens of iterations, render the optimization process challenging,
requiring distinct sets of transformations to bring performance
closest to the machine peak in different problems. For example, the
Burgers problem in Listing~\ref{code:burgers}, given the large number
of arrays accessed, suffers from high register pressure, whereas the
Helmholtz problem in Listing~\ref{code:helmholtz} does not; this
intuitively suggests that the two problems require a different
treatment, based on an in-depth analysis of both data and iteration
spaces. Furthermore, domain knowledge enables transformations that a
general-purpose compiler could not apply, making the optimization
space even larger. In this context, our goal is to understand the
relationship between distinct code transformations, their impact on
cross-loop arithmetic intensity, and to what extent their
composability is effective in a class of problems and architectures.

@@

Data alignment is enforced in two steps. Firstly, all arrays are
allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then,
two dimensional arrays are padded by rounding the number of columns to
the nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the
original size of a basis function array is 3$\times$3 and
$\mbox{\texttt{VL}}=4$ (e.g. AVX processor, with 256-bit long vector
registers and 64-bit double-precision floats). In this case, a padded
version of the array will have size 3$\times$4. The compiler is
explicitly told about data alignment using suitable pragmas; for
example, in the case of the Intel compiler, the annotation
\texttt{$\#$pragma vector aligned} is added before the loop (as shown in
later figures) to inform that all of the memory accesses in the loop
body will be properly aligned. This allows the compiler to issue aligned
load and store instructions, that are notably faster than unaligned
ones.

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-Invariant Code Motion}
\label{sec:code-transf-licm}
From an inspection of the codes in Listings~\ref{code:helmholtz}
and~\ref{code:burgers}, it can be noted that the computation of $A$
involves evaluating many sub-expressions that only depend on two
iteration variables. Since symbols in most of these sub-expressions
are read-only variables, there is ample room for loop-invariant code
motion. Vendor compilers apply this technique, although not in the
systematic way we need for our assembly kernels. We want to overcome
two deficiencies that both Intel and GNU compilers exhibit. First,
they only identify sub-expressions that are invariant with respect to
the innermost loop. This is an issue for sub-expressions depending on
\texttt{i}-\texttt{k}, which are not automatically lifted in the loop
order \texttt{ijk}. Second, the hoisted code is scalar and therefore not
subjected to auto-vectorization.

@@

\KwSty{void} helmholtz(\KwSty{double} A[3][4], \KwSty{double} **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~\KwSty{double} LI$\_$0[4] ALIGN;\\
~~~~\KwSty{double} LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code for the Helmholtz problem in
  Listing~\ref{code:helmholtz} after application of padding, data
  alignment, and loop-invariant code motion, for an AVX architecture. In this
  example, sub-expressions invariant to \texttt{j} are identical to
  those invariant to \texttt{k}, so they can be precomputed once in
  the $r$ loop.}
\label{code:helmholtz-licm}
\end{algorithm}

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
\label{sec:op-vect}
One notable problem of assembly kernels concerns register allocation
and register locality. The critical situation occurs when loop trip
counts and the variables accessed are such that the vector-register
pressure is high. Since the kernel's working set fits the L1 cache, it
is particularly important to optimize register management. Standard
optimizations, such as loop interchange, unroll, and unroll-and-jam,
can be employed to deal with this problem. In COFFEE, these
optimizations are supported either by means of explicit code
transformations (interchange, unroll-and-jam) or indirectly by
delegation to the compiler through standard pragmas (unroll). Tiling
at the level of vector registers is an additional feature of
COFFEE. Based on the observation that the evaluation of the element
matrix can be reduced to a summation of outer products along the
\texttt{j} and \texttt{k} dimensions, a model-driven vector-register
tiling strategy can be implemented. If we consider the code snippet in
Listing~\ref{code:helmholtz-licm} and we ignore the presence of the
operation \texttt{det*W3[i]}, the computation of the element matrix is
abstractly expressible as
\begin{equation}
\scriptsize
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,2
\end{equation}
where $B$ is the set of all basis functions (or temporary variables,
e.g., \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$
are generic problem-dependent subsets. Regardless of the specific
input problem, by abstracting from the presence of all variables
independent of both \texttt{j} and \texttt{k}, the element matrix
computation is always reducible to this form.
Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries
($j,k=0,...,3$) of the element matrix using just 2 vector registers,
that represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B''
\vert = 1$. Values in a register are shuffled each time a product is
performed. Standard compiler auto-vectorization for both GNU and Intel
compilers, instead, executes 4 broadcast operations (i.e., ``splat'' of
a value over all of the register locations) along the outer dimension to
perform the calculation. In addition to incurring a larger number of
cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers
to perform the same 16 evaluations when unroll-and-jam is used, with $f$
being the unroll-and-jam factor.

@@

The compiler applies an ordered sequence of optimization steps to the
ASTs received from FFC. Application of loop-invariant code motion 
(or \emph{licm}) must precede padding
and data alignment, due to the introduction of temporary arrays. These
transformations are always performed because they are likely to improve
the run-time performance, as demonstrated by our results in Section~\ref{sec:analysis-code-transf}.
Based on a cost model (described later), loop interchange, \emph{split},
and \emph{op-vect} may be introduced. Their implementation is centered
on analysis and manipulation of the kernel AST. When \emph{op-vect}
is selected, the compiler outputs AVX or AVX-512 intrinsics code. Any
possible corner cases are handled: for example, if \emph{op-vect} is to
be applied, but the size of the iteration space is not a multiple of the vector
length, then a remainder loop, amenable to auto-vectorization, is inserted.

@@

\subsubsection{Impact of Generalized Loop-Invariant Code Motion}
\label{licm-impact}
